Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2019-08-07 anzjs2018/sections/04_implementation.tex
Update end edit words

The application consists of two components:
the first handles importing static GTFS data and constructing the network,
while the second implements the \rt models and prediction.
We used Rcpp \citep{Rcpp}
to develop the program,
which provides access to R \citep{rcore},
which contains many useful tools for data manipulation and package development,
as well as the speed and memory management capabilities of C++.
The program is implemented in the R package
\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
In this section, we discuss the features of the \rt component
and assess its performance 
both with respect to timing and parameter estimation.

@@

During the development of the application,
we were primarily concerned with ensuring each component of the program
was as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be feasible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provided the memory management control necessary to make this possible.
We also made use of OpenMP for parallelisation,
since each vehicle is modelled independently so it is trivial to scale up without
issues with thread safety.

@@

In each iteration, 
the timing of the various program components was recorded.
Since the number of vehicles travelling at any given time changes throughout the day,
we used the average timings over an off-peak 15~minute window.
Figure~\ref{fig:timings} shows the average timings for 
varying numbers of particles, $N$.

The most time consuming component was ETA writing,
which involved summarising the individual ETAs estimated for each particle.
While not discussed here this involved computing quantiles which uses a sorting algorithm
with complexity $O(N \log N)$,
which explains the slightly quadratic trend in the timings.

The next most intensive step was vehicle updating,
which took about 5~seconds for 8000 particles.
The ETA prediction step only took a few seconds,
although this may increase once a more comprehensive model is developed.
On our 8-core virtual machine the entire process is completed within 20~seconds
when 8000~particles are used.

@@

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\epsilon$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
which led us to use the values $\epsilon \in \{1,2,3,5\}$ for GPS error in meters.
For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to an average vehicle speed variance of between 0.1 and 45~meters per second
over a period of 30~seconds.
For each of combination of these parameter values, including varying $N$,
the following values were calculated:
effective sample size, $N_\text{eff}$;
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the (relative) variance of segment travel time estimates.

@@

Figure~\ref{fig:perf_stats} shows that $N_\text{eff}$ increased
with GPS error and decreased with system noise,
while remaining unaffected by changing $N$.
The degeneration rate decreased with GPS error and increased with $N$,
and was unaffected by system noise.
Large GPS error affects the likelihood,
giving more weight to particles farther from the observation
than does a smaller GPS error.
Conversely, increasing system noise spreads out the particle cloud,
so fewer particles will be near the vehicle.
Thus, these results are not surprising, 
but show that the model is working as expected.

@@

The last measure of performance is the relative variance of segment travel times.
We used relative variance as each segment has a very different travel time distribution,
so to compare the variability between simulations we needed to 
compute the average variance of travel times over all simulations,
taking the average (over road segments) of the ratio between variance for
each individual simulation and variance over all simulations.
If $\tilde z_\ell^{\sigma,\epsilon}$ is a vector of all travel times
along segment $\ell$ during the simulation with GPS error $\sigma$ and system noise $\epsilon$,
then the relative variance $\bar v^{\sigma,\epsilon}$ is calculated by
\begin{equation*}
\begin{split}
v_\ell^{\sigma,\epsilon} &=
\frac{\mathrm{Var}(\tilde z_\ell^{\sigma,\epsilon})}{\mathrm{Var}(\tilde z_\ell)} \\
\bar v^{\sigma,\epsilon} &=
    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\sigma,\epsilon}
\end{split}
\end{equation*}
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
resulted in a higher relative variance,
while simulations with more particles had lower relative variance.
There was no visible effect of system noise on the variability of travel times.

>>>>

The application consists of two components:
the first handles importing static GTFS data and constructing the network,
while the second implements the \rt models and prediction.
We used Rcpp \citep{Rcpp}
to develop the program,
which provides access to R \citep{rcore},
which contains many useful tools for data manipulation and package development,
as well as the speed and memory management capabilities of C++.
The program is implemented in the R package
\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
In this section, we discuss the features of the \rt component
and assess its performance 
both with respect to timing and travel time estimation.

@@

During the development of the application,
we were primarily concerned with ensuring each component of the program
was as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be feasible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provided the memory management control necessary to make this possible.
We also made use of OpenMP for parallelisation,
since each vehicle is modelled independently so it is trivial to scale up without
thread safety concerns.

@@

In each iteration, 
the timings of the various program components were recorded.
Since the number of vehicles travelling at any given time changes throughout the day,
we used the average timings over an off-peak 15~minute window.
Figure~\ref{fig:timings} shows the average timings for 
varying numbers of particles, $N$.

The most time consuming component was ETA writing,
which involved summarising the individual ETAs estimated for each particle.
While not discussed here this involved computing quantiles which uses a sorting algorithm
with complexity $O(N \log N)$,
which explains non-linear relationship with $N$.

The next most intensive step was vehicle updating,
which took about 5~seconds for 8000 particles.
The ETA prediction step only took a few seconds,
although this may increase once a more comprehensive model is developed.
On our 8-core virtual machine,
the average time to process one iteration during off-peak is under 20~seconds
when 8000~particles are used.
Figure~\ref{fig:day-timings} shows iteration timings over the course of a day.

@@

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\omega^2$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
which suggested using values $\omega \in \{1,2,3,5\}$ for GPS error in meters.
For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to an average vehicle speed variation of between 0.1 and 45~meters per second
over a period of 30~seconds.
For each combination of these parameter values, including varying $N$,
the following values were calculated:
effective sample size, $N_\text{eff}$;
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the (relative) variance of segment travel time estimates.

@@

Figure~\ref{fig:perf_stats} shows that $N_\text{eff}$ increases
with GPS error and decreases with system noise,
while remaining unaffected by changing $N$.
The degeneration rate decreases with GPS error and increases with $N$,
but is mostly unaffected by system noise.
Large GPS error affects the likelihood,
giving more weight to particles farther from the observation
than does a smaller GPS error.
Conversely, increasing system noise spreads out the particle cloud,
so fewer particles will be near the vehicle,
decreasing $N_\text{eff}$,
but more plausible trajectories are sampled,
so we see a slight decrease in degeneration rate.
Thus, these results are not surprising, 
but show that the model is working as expected.

@@

The last measure of performance is the relative variance of segment travel times.
We used relative variance as each segment has a very different travel time distribution,
so to compare the variability between simulations we needed to 
compute the average variance of estimated travel times over all simulations,
taking the average (over road segments) of the ratio between variance for
each individual simulation and variance over all simulations.
If $\tilde z_\ell^{\omega,\epsilon}$ is a vector of all travel times
along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\epsilon$,
then the relative variance $\bar v^{\omega,\epsilon}$ is calculated by
\begin{equation*}
\begin{split}
v_\ell^{\omega,\epsilon} &=
\frac{\mathrm{Var}(\tilde z_\ell^{\omega,\epsilon})}{\mathrm{Var}(\tilde z_\ell)} \\
\bar v^{\omega,\epsilon} &=
    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\epsilon}
\end{split}
\end{equation*}
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
resulted in a higher relative variance,
while simulations with more particles had lower relative variance.
There was no visible effect of system noise on the variability of travel times.
